Splitting it up: the \pkg{spduration} split-population duration regression package

### 2015-12-11 manuscript.Rmd
Intro update

Duration models are an important class of statistical estimators that take into account the duration dependency of outcomes we are interested in. For example, the risk of dying is not independent of time. Newborns are at a great risk of dying, but as they grow older this risk declines and then gradually starts to increase again after the age of 9-10. Individual behaviors (smoking, exercise, diet) and structural factors (health care, regulations, urban vs rural) can increase or decrease this risk, but there is an underlying risk that is time dependent and impacts on all units.

However, sometimes we are interested in situations in which not all units are at risk of dying or experiencing some other event of interest. Imagine we are interested in the risk of acquiring the flu. We might think that everyone is at risk of getting the flu, and individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) determine whether individuals catch the flu. If this were the case we could use a standard duration model and estimate how different factors impact on the baseline risk. But there might also be individuals who are immune to the flu, because they received a flu shot, had the same flu virus in the past year, or have some other characteristic that makes it impossible for them to get sick. Given that this is the case we have two underlying populations: An at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions resulting from such a model inaccurate.

>>>>

Duration models are an important class of statistical estimators that take into account the duration dependency of specific outcomes. A prominent example is that the risk of dying depends on the age of an individual. Newborns are at a greater risk of dying, but as they grow older this risk quickly declines and then gradually starts to increase again after the age of 9-10. While individual behavior (smoking, exercising, diet) and structural factors (health care, health regulations, urban vs rural) can increase or decrease the probability of dying over time, the underlying risk is time dependent and impacts on all individuals.

However, there are conditions under which not all individuals have the same underlying risk to experience a specific outcome and might not even be at risk at all. Consider the risk of acquiring a viral infection like the common flu. Let us initially assume that everyone is at risk of infection, and individual behavior (e.g. good hygiene) and structural factors (e.g. workplace) determine whether individuals catch the flu. Under these assumptions a standard duration model and should provide us with efficient estimates of how different behavioral and structural factors impact on the general baseline risk. However, in the overall population there might be individuals that are immune to the flu, because they received vaccination, had the specific flu virus in the past, or have some other characteristic that makes it impossible for them to attract the decease. In this instance we have two underlying populations: An at risk population and an immune one. If the immune population is relatively large, estimates using a standard duration model will be biased and predictions inaccurate.
